\section{Discussion}

Across the analysed datasets, the agreement between TALYS and EXFOR depends strongly on the emitted particle, the incident neutron energy, and the outgoing-energy region considered. An important point is that, in this work, the quantity being changed is the TALYS pre-equilibrium option, \texttt{preeqmode}. Therefore, the largest differences between the calculated spectra are not expected mainly in the low-energy evaporation peak, which is mostly associated with compound-nucleus emission. Instead, the most meaningful differences are expected above the evaporation peak, especially in the intermediate-energy shoulder and in the high-energy tail, where pre-equilibrium emission plays a larger role.

For proton production, the best-performing options are generally \texttt{preeqmode=1} and \texttt{preeqmode=2}. Both of these modes are based on the exciton model, but they differ in how the transition rates are calculated. Mode 1 uses analytical transition rates, while mode 2 uses numerical transition rates and is the default TALYS option. The results show that the default option, \texttt{preeqmode=2}, often gives a reasonable description of the data, especially in the main part of the measured spectrum. However, it is not always clearly the preferred option. In several cases, \texttt{preeqmode=1} gives a slightly better or more stable agreement, particularly when the ratio plot is considered. This means that the default TALYS pre-equilibrium model is a good reference calculation, but it is not universally the best choice for all proton-production cases.

The proton spectra also show why the peak region alone is not sufficient for selecting the best model. In many cases, the different pre-equilibrium modes are very similar near the main peak. This happens because that region is strongly influenced by the compound contribution, which is not the main part being modified when changing \texttt{preeqmode}. The shoulder and high-energy tail are more discriminating because they are more directly connected with pre-equilibrium emission. Therefore, the ratio plots and the component decompositions are essential for deciding between modes 1 and 2, and for identifying cases where modes 3 or 4 deviate more strongly from the experimental data.

For deuteron production, the situation is more complex. The spectra are less smooth than for proton production, and the discrepancies between TALYS and EXFOR are generally larger. In some cases, \texttt{preeqmode=3} gives the best agreement, while in others the selected mode changes with incident energy, including \texttt{preeqmode=4}, \texttt{preeqmode=2}, or \texttt{preeqmode=1}. This shows that there is no single pre-equilibrium option that consistently reproduces all deuteron spectra. Since deuterons are composite particles, their emission is more difficult to describe than proton emission. The results suggest that composite-particle emission may involve stronger contributions from cluster-emission or other non-compound mechanisms, which are not reproduced equally well by all tested TALYS configurations.

A recurring limitation in the deuteron cases is the behaviour of the high-energy tail. TALYS often predicts sharp structures or an abrupt decrease of the cross section at high outgoing energies, while EXFOR still reports non-zero values. This indicates that the fastest emitted deuterons are more difficult to reproduce with the tested models. The improvement observed at higher incident energies suggests that TALYS captures the general trend better in some energy ranges, but the detailed shape of the shoulder and tail remains challenging.

For alpha production, the available comparison is more limited because only one main case was analysed. In this case, \texttt{preeqmode=1} gives the best overall agreement among the tested options. However, the interpretation must be made carefully because the EXFOR data do not strongly constrain the low-\(E_{\mathrm{out}}\) evaporation peak. The mode selection is therefore mainly based on the measured shoulder and tail region. The selected mode reproduces the correct order of magnitude in the experimental range, but the agreement is not uniform: the calculation can overpredict the mid-energy region and then drop too sharply at the high-energy end.

Overall, these results show that there is not one single \texttt{preeqmode} that works best for all reactions and emitted particles. For proton emission, the preferred choices are usually modes 1 and 2, both based on the exciton model, with the default TALYS mode 2 often performing well but not always being the best. For deuteron emission, the preferred mode is more case-dependent, suggesting that composite-particle emission is more sensitive to the details of the model. For alpha emission, mode 1 gives the best result in the analysed case, although the lack of low-energy EXFOR points limits the strength of the conclusion.

The main methodological conclusion is that model assessment should not rely only on visual agreement near the spectral peak. The most informative regions are the shoulder and high-energy tail, because these are the regions where pre-equilibrium emission has the strongest effect. The ratio plots make systematic overprediction and underprediction easier to identify, while the component decompositions clarify which reaction mechanism dominates each part of the spectrum. Combining these diagnostics gives a more reliable basis for selecting the best TALYS configuration than the total spectra alone.

