\section{Methodology}

In this work, TALYS simulations are compared to experimental energy spectra, and we evaluate the
\emph{sensitivity} of the calculated spectra to model choices by identifying which settings provide the
best agreement for each case. The following reactions are studied:
\begin{itemize}
    \item \textbf{Proton production:} $(n,px)$ on several targets (Si, Fe, Cr, V, Co), at incident neutron energies spanning
    approximately 14--60~MeV depending on the dataset;
    \item \textbf{Deuteron production:} $(n,dx)$ on Si for selected energies;
    \item \textbf{Alpha production:} $(n,\alpha x)$ on Si for selected energies.
\end{itemize}

To investigate sensitivity to the modelling of the intermediate stage, four TALYS calculations were
performed for each reaction case, corresponding to four pre-equilibrium configurations (\texttt{preeq1},
\texttt{preeq2}, \texttt{preeq3} and \texttt{preeq4}), while keeping the other inputs fixed. This ensures that
differences between \texttt{preeq1--preeq4} can be attributed primarily to the treatment of pre-equilibrium
dynamics and its competition with compound emission. At intermediate incident energies, the outgoing
light-ion spectrum is typically a superposition of direct/direct-like emission, pre-equilibrium emission,
and compound-nucleus evaporation. \Cref{tab:preeq_settings} summarizes the four configurations used in this project.

\begin{table}[ht]
\centering
\small
\setlength{\tabcolsep}{6pt}
\renewcommand{\arraystretch}{1.15}
\begin{tabularx}{\textwidth}{@{} l X  @{}}
\toprule
Configuration & Expected qualitative effect on spectra \\
\midrule
\texttt{preeq1} &
Baseline option; primarily affects the shoulder and high-energy tail. \\

\texttt{preeq2} &
Reference configuration used for most EXFOR comparisons; sets the nominal pre-equilibrium strength and slope. \\

\texttt{preeq3} &
Modified pre-equilibrium dynamics, e.g.\ transition rates, exciton options, or angular systematics, which can change spectral hardness and overall normalisation. \\

\texttt{preeq4} &
Enhanced treatment, often including multiple pre-equilibrium emission and/or cluster-emission options; can increase yield at intermediate-to-high $E_{\mathrm{out}}$ and reshape d/$\alpha$ spectra. \\
\bottomrule
\end{tabularx}
\caption{Definition of the four TALYS configurations used in this work.}
\label{tab:preeq_settings}
\end{table}

In this report, the four modes are treated mainly as different ways of testing the sensitivity of the spectra to the pre-equilibrium stage. In general, the modes affect mostly the intermediate- and high-energy regions of the spectra, which means that the shoulder and the tail are especially important for deciding which mode works better. For composite particles, such as deuterons and alpha particles, the effect of the mode choice can be even more visible since it usually includes stronger direct-like or cluster-emission contributions.


In TALYS, pre-equilibrium emission for nucleon-induced reactions is commonly described within an
exciton-model framework, in which the excited nucleus is represented by a particle--hole configuration.
The exciton number
\begin{equation}
n = p + h,
\label{eq:excitons}
\end{equation}
counts the number of excited particles $p$ above the Fermi surface and holes $h$ below it. Starting from
an initial configuration produced by the first interaction(s), the system evolves through two-body
residual interactions toward more complex configurations, meaning larger $n$, while emission competes with this
equilibration process. As a result, pre-equilibrium emission tends to populate the intermediate-to-high
outgoing-energy region of the spectra compared to compound evaporation.

To satisfy the main objective of this project---comparing TALYS outgoing-particle spectra against EXFOR and identifying which experimental datasets and incident neutron energies are most informative for constraining TALYS predictions---a consistent comparison methodology was applied at each incident neutron energy and for each pre-equilibrium option (\texttt{preeqmode}=1--4). For every case, the four TALYS total spectra were compared with the EXFOR spectrum to assess the peak height and position, mid-energy shoulder behaviour, and the high-energy tail. In addition, the TALYS mechanism components were plotted together with EXFOR to identify how each component contributes to the overall spectrum.

To complement this qualitative assessment with a quantitative criterion, TALYS totals were evaluated at the EXFOR outgoing-energy points by interpolation, and a point-by-point ratio was computed as
\[
R(E_{\mathrm{out}})=\frac{\mathrm{TALYS}(E_{\mathrm{out}})}{\mathrm{EXFOR}(E_{\mathrm{out}})}.
\]
The ratio plots highlight systematic biases, over- or underprediction, and make the magnitude of deviations easier to judge than the totals alone. A ratio close to
\[
R(E_{\mathrm{out}})=1
\]
means that TALYS and EXFOR are in good agreement at that outgoing energy. Values above 1 indicate that TALYS overpredicts the experimental data, while values below 1 indicate an underprediction. Finally, based on the totals comparison and the ratio behaviour, a \texttt{best preeq mode} was selected for each case.

The new figures used below contain, in a single diagnostic plot, the linear comparison of totals and EXFOR, the same comparison in logarithmic scale, the ratio plot for all pre-equilibrium modes, and the spectral-component decomposition for the selected mode.

In the logarithmic panels, special care was taken. In some cases, the TALYS curves fall to extremely small values close to the numerical plotting limit. These values do not correspond to a physically meaningful part of the comparison and can make the relevant differences between modes harder to see. For this reason, the figures were adjusted to focus on the energy and cross-section ranges that have physical meaning.
